FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-06-17 paper.tex
Edit primer on floating point

>>>>

\Fp{} numbers are a way of representing real numbers in a computer.
However, since we limit our precision to a fixed, finite set of bits, \fp{} values are \emph{inherently approximate} and necessarily incur some error during computation.
This is not bad; it is a trade-off: we get fast and compact representations that are good enough for most use cases.
However, the downside is that \fp{} numbers \emph{do not} behave like real numbers.

% FIXME: more resources!
There are many excellent resources\cite{knuthArtComputerProgramming1997,torontoPracticallyAccurateFloatingPoint2014} on understanding the theory behind \fp{};
we will explain the most essential aspects here briefly.

A \fp value approximates real number $r \in \mathbb{R}$ by expressing $r$ in scientific notation $r = s \times f \times b^{e-q}$,
for some base $b$,
an integer exponent $e \in \mathbb{N}$,
a fixed value called an \emph{excess} $q$,\footnote{The purpose of the excess $q$ is so that with value of $e > q$ we can represent positive exponents, and $e < q$ we can represent negative exponents.}
a \emph{sign} $s \in \{-1, 1\}$,
and a \emph{fraction} $f$ such that $0 \leq f < b$.\footnote{Another name for the fraction part commonly seen in the literature is ``mantissa'' or ``significand''.
  You might occasionally also see ``characteristic'' referring to the exponent.
  We will stick to ``fraction'' and ``exponent'' however.}

We pack $s$, $e$, and $f$ into bit fields.
This gives us a compact representation for both very small and very large values: because the exponent is variable, we can represent numbers as big as Avogadro's number or as small as Planck's constant with the same number of bits.
However, since we have only a finite number of bits $p$, we have to round $f$ and the range of $e$ is limited.
Moreover, there is some error between the real number $r$ and the value we can express with a float.
This is why $0.1 + 0.2 = 0.30000000000000004$---$0.3$ is not precisely representable.

To make matters worse, things get tricky when we do math.
Adding e.g. Avogadro's number and Planck's constant together is beyond 16-, 32-, and 64-bit floats to represent.
In fact, whenever the exponents differ by more than $p+2$, the result of the addition will just be the same as the larger number.
In the case of Avogadro's number, adding any value less than $FIXME$ will just result in Avogadro's number again.

The gradual accumulation of these errors and other representation issues leads to surprising results!
% FIXME: more examples of a NaN popping up out of nowhere, e.g. with bad addition
